---
layout: page
title: R for reproducible scientific analysis
subtitle: Manipulating data.frames
minutes: 130
---

```{r, include=FALSE}
source("tools/chunk-options.R")
```

> ## Learning objectives {.objectives}
>
> - Be able to use the six major dplyr verbs (`filter`, `select`, `arrange`, `mutate`, `group_by`, `summarize`)
> - Be able to use and understand the advantages of the `magrittr` pipe: `%>%`
>

It is an often bemoaned fact that a data scientist spends much, and often most, of her time wrangling data: getting it organized and clean. In this lesson we will learn an efficient set of tools that can handle the vast majority of most data management tasks. 

Enter `dplyr`, a package for making data manipulation easier. More on `dplyr` later. `dplyr` is part of `tidyverse`, so it is already installed on your machine. You can load it individually, or with the other tidyverse packages like this:

```{r, warning=FALSE}
library(tidyverse)
```

Those messages and conflicts are normal. The conflicts are R telling you that there are two packages with functions named "filter" and "lag". When R gives you red text, it's not always a bad thing, but it does mean you should pay attention and try to understand what it's trying to tell you.

Remember that you only have to install each package once (per computer), but you have to load them for each R session in which you want to use them.

You also have to load any data you want to use each time you start a new R session. So, if it's not already loaded, read in the gapminder data. We're going to use tidyverse's `read_csv` instead of base R's `read.csv` here. It has a few nice features; the most obvious is that it makes a special kind of data.frame that only prints the first ten rows instead of all 1704.

```{r load, warning = FALSE, message = FALSE}
gapminder <- read_csv('data/gapminder-FiveYearData.csv')
class(gapminder)
gapminder
```

You can always convert a data.frame into this special kind of data.frame like this:

```{r}
gapminder <- tbl_df(gapminder)
```

## What is dplyr?

The package `dplyr` is a fairly new (2014) package that tries to provide easy
tools for the most common data manipulation tasks. It is built to work directly
with data frames. The thinking behind it was largely inspired by the package
`plyr` which has been in use for some time but suffered from being slow in some
cases.` dplyr` addresses this by porting much of the computation to C++. An
additional feature is the ability to work with data stored directly in an
external database. The benefits of doing this are that the data can be managed
natively in a relational database, queries can be conducted on that database,
and only the results of the query returned.

This addresses a common problem with R in that all operations are conducted in
memory and thus the amount of data you can work with is limited by available
memory. The database connections essentially remove that limitation in that you
can have a database of many 100s GB, conduct queries on it directly and pull
back just what you need for analysis in R.


### The five tasks of `dplyr`

There are five actions we often want to apply to a tabular dataset:

- Filter rows
- Filter columns
- Arrange rows
- Make new columns
- Summarize groups

We are about to see how to do each of those things using the `dplyr` package. Everything we're going to learn to do can also be done using "base R", but `dplyr` makes it easier, and the syntax is consistent, and it actually makes the computations faster. 

#### `filter()`

Suppose we want to see just the gapminder data for the USA. First, we need to know how "USA" is written in the dataset: Is it USA or United States or what? We can see all the unique values of a variable with the `unique` function.

```{r unique}
unique(gapminder$country)
```

Okay, now we want to see just the rows of the data.frame where country is "United States". The syntax for all `dplyr` functions is the same: The first argument is the data.frame, the rest of the arguments are whatever you want to do in that data.frame. 

```{r filter}
filter(gapminder, country == "United States")
```

We can also apply multiple conditions, e.g. the US after 2000:

```{r filter and}
filter(gapminder, country == "United States" & year > 2000)
```

We can also use "or" conditions with the vertical pipe: `|`. Notice that the variable (column) names don't go in quotes, but values of character variables do. 

```{r filter or}
filter(gapminder, country == "United States" | country == "Mexico")
```

A good, handy reference list for the operators (and, or, etc) can be found [here](http://www.statmethods.net/management/operators.html).

#### `select()`

`filter` returned a subset of the data.frame's rows. `select` returns a subset of the data.frame's columns.

Suppose we only want to see country and life expectancy. 

```{r select, eval = FALSE}
select(gapminder, country, lifeExp)
```

We can choose which columns we don't want

```{r}
select(gapminder, -continent, income = gdpPercap)
```

And we can rename columns

```{r}
select(gapminder, ThePlace = country, HowLongTheyLive = lifeExp)
```

As usual, R isn't saving any of these outputs; just printing them to the screen. If we want to keep them around, we need to assign them to a variable.

```{r}
justUS = filter(gapminder, country == "United States")
USdata = select(justUS, -country, -continent)
USdata
```


> #### Subsetting {.challenge}
>
> - Subset the gapminder data to only Oceania countries post-1980.
> - Remove the continent column
> - Make a scatter plot of gdpPercap vs. population colored by country
>
> **Advanced** How would you determine the median population for the North American countries between 1970 and 1980?
>
> **Bonus** This can be done using base R's subsetting, but this class doesn't teach how. Do the original challenge without the `filter` and `select` functions. Feel free to consult Google, helpfiles, etc. to figure out how.
> 


#### `arrange()`

You can order the rows of a data.frame by a variable using `arrange`. Suppose we want to see the most populous countries: 

```{r arrange}
arrange(gapminder, pop)
```

Hmm, we didn't get the most populous countries. By default, `arrange` sorts the variable in *increasing* order. We could see the most populous countries by examining the `tail` of the last command, or we can sort the data.frame by descending population by wrapping the variable in `desc()`:

```{r arrange desc}
arrange(gapminder, desc(pop))
```

`arrange` can also sort by multiple variables. It will sort the data.frame by the first variable, and if there are any ties in that variable, they will be sorted by the next variable, and so on. Here we sort from newest to oldest, and within year from richest to poorest:

```{r arrange multi}
arrange(gapminder, desc(year), desc(gdpPercap))
```

**Shoutout Q: Would we get the same output if we switched the order of `desc(year)` and `desc(gdpPercap)` in the last line?**


#### `mutate()`

We have learned how to drop rows, drop columns, and rearrange rows. To make a new column we use the `mutate` function. As usual, the first argument is a data.frame. The second argument is the name of the new column you want to create, followed by an equal sign, followed by what to put in that column. You can reference other variables in the data.frame, and `mutate` will treat each row independently. E.g. we can calculate the total GDP of each country in each year by multiplying the per-capita GDP by the population. 

```{r mutate}
mutate(gapminder, total_gdp = gdpPercap * pop)
```

**Shoutout Q: How would we view the highest-total-gdp countries?**

Note that didn't change gapminder: We didn't assign the output to anything, so it was just printed, with the new column. If we want to modify our gapminder data.frame, we can assign the output of `mutate` back to the gapminder variable, but be careful doing this -- if you make a mistake, you can't just re-run that line of code, you'll need to go back to loading the gapminder data.frame.

Also, you can create multiple columns in one call to `mutate`, even using variables that you just created, separating them with commas:

```{r}
gapminder = mutate(gapminder, 
                   total_gdp = gdpPercap * pop,
                   log_gdp = log10(total_gdp))
```


> #### MCQ: Data Reduction {.challenge}
>
> Produce a data.frame with only the names, years, and per-capita GDP of countries where per capita gdp is less than a dollar a day sorted from most- to least-recent.
>
> - Tip: The `gdpPercap` variable is annual gdp. You'll need to adjust.
> - Tip: For complex tasks, it often helps to use pencil and paper to write/draw/map the various steps needed and how they fit together before writing any code.
> 
> What is the annual per-capita gdp, rounded to the nearest dollar, of the first row in the data.frame?
>
> a. $278
> b. $312
> c. $331
> d. $339
>
> **Advanced**: Use dplyr functions and ggplot to plot per-capita GDP versus population for North American countries after 1970.
> - Once you've made the graph, transform both axes to a log10 scale. There are two ways to do this, one by creating new columns in the data frame, and another using functions provided by ggplot to transform the axes. Implement both, in that order. Which do you prefer and why?
>



#### C'est ne pas une pipe

Suppose we want to look at all the countries where life expectancy is greater than 80 years, sorted from poorest to richest. First, we `filter`, then we `arrange`. We could assign the intermediate data.frame to a variable:

```{r filter and arrange assign}
lifeExpGreater80 = filter(gapminder, lifeExp > 80)
(lifeExpGreater80sorted = arrange(lifeExpGreater80, gdpPercap))
```

In this case it doesn't much matter, but we make a whole new data.frame (`lifeExpGreater80`) and only use it once; that's a little wasteful of system resources, and it clutters our environment. If the data are large, that can be a big problem. 

Or, we could nest each function so that it appears on one line:

```{r filter and arrange assign oneline}
arrange(filter(gapminder, lifeExp > 80), gdpPercap)
```

This would become difficult to read if we are performing a number of operations that would require a repeated nesting. But...

There is a better way, and it makes both writing and reading the code easier. The pipe from the `magrittr` package (which is automatically installed and loaded with `dplyr` and `tidyverse`) takes the output of first line, and plugs it in as the first argument of the next line. Since many `tidyverse` functions expect a data.frame as the first argument and output a data.frame, this works fluidly.

```{r filter and arrange pipe}
filter(gapminder, lifeExp > 80) %>%
    arrange(gdpPercap)
```

To demonstrate how it works, here are some examples where it's unnecessary. 

```{r pipe simple}
4 %>% sqrt()
2 ^ 2 %>% sum(1)
```

Whatever goes through the pipe becomes the first argument of the function after the pipe. This is convenient, because all `dplyr` functions produce a data.frame as their output and take a data.frame as the first argument. Since R ignores white-space, we can put each function on a new line, which RStudio will automatically indent, making everything easy to read. Now each line represents a step in a sequential operation. You can read this as "Take the gapminder data.frame, filter to the rows where lifeExp is greater than 80, and arrange by gdpPercap." 

```{r pipe}
gapminder %>%
    filter(lifeExp > 80) %>%
    arrange(gdpPercap)
```

Making your code easier for humans to read will save you lots of time. The human reading it is usually future-you, and operations that seem simple when you're writing them will look like gibberish when you're three weeks removed from them, let alone three months or three years or another person. Make your code as easy to read as possible by using the pipe where appropriate, leaving white space, using descriptive variable names, being consistent with spacing and naming, and liberally commenting code.

> #### Challenge: Data Reduction with Pipes {.challenge}
>
> Copy the code you (or the instructor) wrote to solve the previous MCQ Data Reduction challenge. Rewrite it using pipes (i.e. no assignment and no nested functions)
>
>


#### `summarize()`

Often we want to calculate a new variable, but rather than keeping each row as an independent observation, we want to group observations together to calculate some summary statistic. To do this we need two functions, one to do the grouping and one to calculate the summary statistic: `group_by` and `summarize`. By itself `group_by` doesn't change a data.frame; it just sets up the grouping. `summarize` then goes over each group in the data.frame and does whatever calculation you want. E.g. suppose we want the average global gdp for each year. While we're at it, let's calculate the mean and median and see how they differ. 

```{r summarize}
gapminder %>%
    group_by(year) %>%
    summarize(mean_gdp = mean(gdpPercap), median_gdp = median(gdpPercap))
```

**Shoutout Q: Note that `summarize` eliminates any other columns. Why? What else can it do? E.g. What country should it list for the year 1952!?**

There are several different summary statistics that can be generated from our data. The R base package provides many built-in functions such as `mean`, `median`, `min`, `max`, and `range`.  By default, all **R functions operating on vectors that contains missing data will return NA**. It's a way to make sure that users know they have missing data, and make a conscious decision on how to deal with it. When dealing with simple statistics like the mean, the easiest way to ignore `NA` (the missing data) is to use `na.rm=TRUE` (`rm` stands for remove). An alternate option is to use the function `is.na()`, which evaluates to true if the value passed to it is not a number. This function is more useful as a part of a filter, where you can filter out everything that is not a number. For that purpose you would do something like

```{r isna, purl = FALSE, eval=FALSE}
gapminder %>%
  filter(!is.na(someColumn)) 
```

The `!` symbol negates it, so we're asking for everything that is not an `NA`. 


We often want to calculate the number of entries within a group. E.g. we might wonder if our dataset is balanced by country. We can do this with the `n()` function, or `dplyr` provides a `count` function as a convenience:

```{r count}
gapminder %>%
    group_by(country) %>%
    summarize(number_entries = n())
count(gapminder, country)
```


We can also do multiple groupings. Suppose we want the maximum life expectancy in each continent for each year. We group by continent and year and calculate the maximum with the `max` function:

```{r multiple groups}
gapminder %>%
    group_by(continent, year) %>%
    summarize(longest_life = max(lifeExp))
```

Hmm, we got the longest life expectancy for each continent-year, but we didn't get the country. To get the country, we have to ask R "Where lifeExp is at a maximum, what is the entry in country?" For that we use the `which.max` function. `max` returns the maximum value; `which.max` returns the location of the maximum value.

```{r which.max}
max(c(1, 7, 4))
which.max(c(1, 7, 4))
```

Now, back to the question: Where lifeExp is at a maximum, what is the entry in country? 

```{r which.max lifeExp}
gapminder %>%
    group_by(continent, year) %>%
    summarize(longest_life = max(lifeExp), country = country[which.max(lifeExp)])
```




> #### Challenge -- Part 1 {.challenge}
>
> - Calculate a new column: the total GDP of each country in each year. 
> - Calculate the variance -- `var()` of countries' gdps in each year.
> - Is country-level GDP getting more or less equal over time?
> - Plot your findings.
> 

> #### Challenge -- Part 2 {.challenge}
> 
> - Modify the code you just wrote to calculate the variance in both country-level GDP and per-capita GDP.
> - Do both measures support the conclusion you arrived at above?
> 

> #### Challenge -- Part 3 (Advanced) {.challenge}
> 
> The above plotting exercise asked you to plot summarized information, but it is generally preferable to avoid summarizing before plotting. Can you generate a plot that shows the information you calculated in Part 1 without summarizing?  
>
> - Hint: `ggplot` interprets the `gapminder$year` as a numeric variable, which may be okay, but there are some plot types for which you need `ggplot` to see `gapminder$year` as a category. You can accomplish this by wrapping it in `factor` -- e.g. `ggplot(gapminder, aes(x = factor(year) ...`
>

#### Resources

That is the core of `dplyr`'s functionality, but it does more. RStudio makes a great [cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf) that covers all the `dplyr` functions we just learned, plus what we will learn in the next lesson: keeping data tidy.


## Challenge solutions

> #### Solution to challenge Subsetting {.challenge}
>
> - Subset the gapminder data to only Oceania countries post-1980.
> - Remove the continent column
> - Make a scatter plot of gdpPercap vs. population colored by country
>
> ```{r challenge subsetting a}
> oc1980 = filter(gapminder, continent == "Oceania" & year > 1980)
> oc1980less = select(oc1980, -continent)
> library('ggplot2')
> ggplot(oc1980less, aes(x = gdpPercap, y = lifeExp, color = country)) +
>   geom_point()
> ```
>
> **Advanced** How would you determine the median population for the North American countries between 1970 and 1980?
>
> ```{r challenge subsetting b}
> noAm = filter(gapminder, country == "United States" | 
>                 country == "Canada" | country == "Mexico" | 
>                 country == "Puerto Rico" & (year > 1970 & year < 1980)
>               )
> noAmPop = select(noAm, pop)
> median(noAmPop)
> noAmPop
> as.integer(noAmPop)
> median(unlist(noAmPop))
> ```
>
> **Bonus** This can be done using base R's subsetting, but this class doesn't teach how. Do the original challenge without the `filter` and `select` functions. Feel free to consult Google, helpfiles, etc. to figure out how.
> 
> ```{r challenge subsetting c}
> noAm2 = gapminder[(gapminder$country == "United States") |
>                     (gapminder$country == "Mexico") |
>                     (gapminder$country == "Canada") |
>                     (gapminder$country == "Puerto Rico") &
>                     ((gapminder$year > 1970) &
>                     (gapminder$year < 1980)),]
> median(noAm2$pop)
> ```
>

> #### Solution to challenge MCQ: Data Reduction {.challenge}
>
> Produce a data.frame with only the names, years, and per-capita GDP of countries where per capita gdp is less than a dollar a day sorted from most- to least-recent.
>
> - Tip: The `gdpPercap` variable is annual gdp. You'll need to adjust.
> - Tip: For complex tasks, it often helps to use pencil and paper to write/draw/map the various steps needed and how they fit together before writing any code.
> 
> What is the annual per-capita gdp, rounded to the nearest dollar, of the first row in the data.frame?
>
> a. $278
> b. $312
> c. $331
> d. $339
> 
> ```{r data reduction A}
> dailyGDP = mutate(gapminder, onedayGDP = gdpPercap / 365)
> dailyGDP = filter(dailyGDP, onedayGDP < 1)
> dailyGDP = select(dailyGDP, country, year, gdpPercap)
> dailyGDP[1,]
> ```
>
> **Advanced**: Use dplyr functions and ggplot to plot per-capita GDP versus population for North American countries after 1970.
> - Once you've made the graph, transform both axes to a log10 scale. There are two ways to do this, one by creating new columns in the data frame, and another using functions provided by ggplot to transform the axes. Implement both, in that order. Which do you prefer and why?
>
> ```{r data reduction B}
> noAm = filter(gapminder, country == "United States" | 
>                 country == "Canada" | country == "Mexico" | 
>                 country == "Puerto Rico" & year > 1970
>               )
> ggplot(noAm, aes(x = gdpPercap, y = pop, color = country)) +
>   geom_point() + 
>   scale_x_log10() +
>   scale_y_log10()
> # OR
> noAmlog10 = mutate(noAm, log10pop = log10(pop),
>                    log10gdp = log10(gdpPercap))
> ggplot(noAmlog10, aes(x = log10gdp, y = log10pop, color = country)) +
>   geom_point()
> ```
>

> #### Challenge: Data Reduction with Pipes {.challenge}
>
> Copy the code you (or the instructor) wrote to solve the previous MCQ Data Reduction challenge. Rewrite it using pipes (i.e. no assignment and no nested functions)
>
> ```{r data reduction pipes}
> # previous challenge with pipes
> dailyGDP = mutate(gapminder, onedayGDP = gdpPercap / 365)
> dailyGDP = filter(dailyGDP, onedayGDP < 1)
> dailyGDP = select(dailyGDP, country, year, gdpPercap)
> # BECOMES
> smallGDP = gapminder %>%
>   mutate(onedayGDP = gdpPercap / 365) %>%
>   filter(onedayGDP < 1) %>%
>   select(country, year, gdpPercap)
> smallGDP[1,]
> # OR, more fancy (without an intermediate temp variable)
> (gapminder %>%
>   mutate(onedayGDP = gdpPercap / 365) %>%
>   filter(onedayGDP < 1) %>%
>   select(country, year, gdpPercap))[1,]
> ```
>

> #### Challenge -- Part 1 {.challenge}
>
> - Calculate a new column: the total GDP of each country in each year. 
> - Calculate the variance -- `var()` of countries' gdps in each year.
> - Is country-level GDP getting more or less equal over time?
> - Plot your findings.
> 
> ```{r challenge LessonEnd part1}
> varGDP = gapminder %>%
>   mutate(totalGDP = gdpPercap * pop) %>%
>   group_by(year) %>%
>   summarize(varTotGDP = var(totalGDP))
> ggplot(varGDP, aes(x = year, y = varTotGDP)) +
>   geom_point()
> ```
>

> #### Challenge -- Part 2 {.challenge}
> 
> - Modify the code you just wrote to calculate the variance in both country-level GDP and per-capita GDP.
> - Do both measures support the conclusion you arrived at above?
> 
> ```{r challenge LessonEnd part2}
> varGDP = gapminder %>%
>   mutate(totalGDP = gdpPercap * pop) %>%
>   group_by(year) %>%
>   summarize(varTotGDP = var(totalGDP),
>             varPerCapGDP = var(gdpPercap)
>             )
> ggplot(varGDP) +
>   geom_point(color = "red", aes(x = year, y = varTotGDP)) +
>   geom_point(color = "blue", aes(x = year, y = varPerCapGDP))
> ```

> #### Challenge -- Part 3 (Advanced) {.challenge}
> 
> The above plotting exercise asked you to plot summarized information, but it is generally preferable to avoid summarizing before plotting. Can you generate a plot that shows the information you calculated in Part 1 without summarizing?  
>
> - Hint: `ggplot` interprets the `gapminder$year` as a numeric variable, which may be okay, but there are some plot types for which you need `ggplot` to see `gapminder$year` as a category. You can accomplish this by wrapping it in `factor` -- e.g. `ggplot(gapminder, aes(x = factor(year) ...`
>
>
> ```{r challenge LessonEnd part3, fig.width=8}
> gapminder %>%
>   mutate(totalGDP = gdpPercap * pop) %>%
> ggplot(aes(x = factor(year), y = totalGDP)) +
>   geom_violin() +
>   scale_y_log10() 
>  ```
>