---
layout: page
title: R for reproducible scientific analysis
subtitle: Statistical Models
minutes: 30
---

```{r setup, include = FALSE}
library(tidyverse)
source("tools/chunk-options.R")
gapminder <- read_csv('data/gapminder-FiveYearData.csv')
```

> ## Learning Objectives {.objectives}
>
> * Understand how to execute and interpret basic statistical models
> * Learn to querry and extract from lists
> * Use `broom` to work with models

### Linear models

This workshop can't and won't teach you statistical modeling, but here is the syntax you need to know to use R's basic statistical modeling infrastructure.

`lm` is the function for a linear model. `lm` expects a formula as its first argument. Formulas in R are specified with a tilde separating the left and right hand sides: `DV ~ IV1 + IV2 + ...`. The second argument to `lm` is the name of the data.frame in which the variables are to be found. For example, to model life expectancy as a function of gdp:

```{r}
lm(lifeExp ~ gdpPercap, gapminder)
```

We can include additional predictors by separating them with a `+`. Now we will assign the results of the model to a variable called `model` and then get a more detailed description of the results by calling the `summary` function.

```{r}
model <- lm(lifeExp ~ gdpPercap + year, gapminder)
summary(model)
```

Notice that the same `summary` function gives summary information of a different type depending on whether its argument is a data.frame, a linear model, or something else. That's handy.

We can specify interaction effects by separating variables with `*`:

```{r}
interaction_model <- lm(lifeExp ~ gdpPercap * continent, gapminder)
summary(interaction_model)
```



### `broom`

The package `broom` is another `tidyverse`-family member. It is built to make working with models easier. Since you won't always be working with models, it doesn't load automatically with `tidyverse` (but does install with it), so we load it with library.

```{r}
library(broom)
```

We can get a nice data.frame output of the model summary.

```{r}
tidy(model)
```

We can create a data.frame that has the data that went into the model plus a bunch of new columns based on the model, such as residuals and predicted values. Note that all the newly added columns' names start with ".".

```{r}
modelOut = augment(model)
head(modelOut)
```

This is useful for checking model assumptions, looking for anomalous points that may indicate omitted variables, etc. For example, it looks like our model underpredicts short life expectancies:

```{r}
ggplot(modelOut, aes(lifeExp, .resid, color = year)) + 
  geom_point()
```


### glm and beyond

Finally, the specification of generalized linear models such as logistic or Poisson regressions is very similar via the `glm` command. See `?glm` and the web for help. Beyond glm's, the statistical capabilities of R are extensive. A Google search for whatever you are interested in will get you started.



> ## Challenge - A plot and a model {.challenge}
>
> - Make a scatterplot of gdpPercap versus year.  
> - Add a smoother and specify `method = lm` to get a linear fit.  
> - Run a linear regression of gpdPercap on year and use `tidy` to extract the model results.  
> - Do your plot and model point to the same conclusions? Which do you find easier to interpret?
>
> Advanced  
>
> - Does the change in gdpPercap over time vary across continents?  
>   - Hint: An interaction model can answer that question.


> ## Alternatve challenge - Stock prices {.challenge}
>
> - Using the stock data you tidy'd earlier, fit a simple linear model of stock performance.   
> - Extract the model coefficients into a data.frame.  
> - Fortify the data with residuals, predicted values, etc.  
> - Examine (however you wish) residuals by stock. Is the model particularly over or underpredicting any particular stock? How could you improve the model?  
> - **Advanced**: Build that better model.