---
layout: page
title: R for reproducible scientific analysis
subtitle: Statistical Models
minutes: 30
---

```{r setup, include = FALSE}
source("tools/chunk-options.R")
gapminder <- read.csv('data/gapminder-FiveYearData.csv')
```

> ## Learning Objectives {.objectives}
>
> * Understand how to execute and interpret basic statistical models. 
> 

### Linear models

This workshop can't and won't teach you statistical modeling, but here is the syntax you need to know to use R's basic statistical modeling infrastructure.

`lm` is the function for a linear model. `lm` expects a formula as its first argument. Formulas in R are specified with a tilde separating the left and right hand sides: `DV ~ IV1 + IV2 + ...`. The second argument to `lm` is the name of the data.frame in which the variables are to be found. For example, to model life expectancy as a function of income:

```{r}
lm(lifeExp ~ gdpPercap, gapminder)
```

We can include additional predictors by separating them with a `+`. Now we will assign the results of the model to a variable called `model` and then get a more detailed description of the results by calling the `summary` function.

```{r}
model <- lm(lifeExp ~ gdpPercap + year, gapminder)
summary(model)
```

Notice that the same `summary` function gives summary information of a different type depending on whether its argument is a data.frame, a linear model, or something else. That's handy.

We can specify interaction effects by separating variables with `*`:

```{r}
model2 <- lm(lifeExp ~ gdpPercap * continent, gapminder)
summary(model2)
```

### glm and beyond

Finally, the specification of generalized linear models such as logistic or Poisson regressions is very similar via the `glm` command. See `?glm` and the web for help. Beyond glm's, the statistical capabilities of R are extensive. A Google search for whatever you are interested in will get you started.



> ## Challenge - A plot and a model {.challenge}
>
> - Make a scatterplot of gdpPercap versus year.
> - Add a smoother and specify `method = lm` to get a linear fit.
> - Run a linear regression of gpdPercap on year.
> - Do your plot and model point to the same conclusions? Which do you find easier to interpret?
>