---
title: "Introduction to RMarkdown"
author: "Michael Koontz"
date: "10/5/2016"
output: html_document
---

<!--- Start by opening a new RMarkdown file and deleting everything after the header. Knit the document to start ---!>

This is an RMarkdown file, which combines text and executed R code into a single, reproducible document. First, we write the text and code that we want to execute, then we 'knit' those pieces together, which creates a single document (in our case an .html file) as a finished product.

We write plain text in the white space within this .Rmd file.

If we want to write R code, and have that R code executed, then we use the following syntax in the .Rmd file to create a 'chunk' of R code. Three back ticks, open brace, a lower-case 'r', closed curly brace, the code we want to execute, and then three back ticks at the end. Notice the area between the back ticks turns grey.

```{r}
4 + 5
```

These back ticks and curly braces disappear in the rendered document, because they have a specific meaning within RMarkdown.

Here's a more relevant example:

```{r nameOfChunk}
str(trees) # The structure of the built-in dataset trees
```

<!--- Exercise --->
Open up a new markdown file


You can also format the plain text using the markdown language. Here's an external link to a great cheat sheet: <https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf>

# Headers! ('#' does not designate a comment in the markdown language)

<!---  this is a comment in RMarkdown, with a start AND an end component ---> 


## A smaller header
##### The smallest of headers.

### Italics and bold font
Italics using the asterisk. I am *really* emphasizing this word. Bold text using pairs of asterisks. Here is another way to **really** make sure my wording is strong.

### Inline R Code
We can also use R code nested within our text. If we want to output the mean value of the Girth column in the trees dataset, rounded off to 2 decimal places, we can say: The mean girth of all trees is `r round(x = mean(trees$Girth), digits = 2)`. And then we can keep typing!

### Greek letters, math symbols, and equations

Inline math like this $4+4$. Or an inline Greek symbol, let's say $\alpha$.

Here's a linear model inline: $y_i = \beta_0 + \beta_1x_i + \epsilon$

To use TeX formatting and make multiple lined equations (these are also aligned by the equal sign and "is distributed as" tilde).

$$
\begin{aligned}
y_i &= \beta_{0} + \beta_1x_i + \epsilon \\
\epsilon &\sim Normal(0, \sigma^2)
\end{aligned}
$$

```{r readTheDataNoShow, echo = FALSE}
library(babynames)
```

The next chunk will set the eval= argument to FALSE so that the code is printed to the knitted document, but not executed in R.

```{r noEvaluationChunk, eval = FALSE}
str(babynames)
```

### Typewriter font.

Use the code font to write text in typewriter font with grey background.

Here I'm typing plain text, but I want to emphasize that the `mean()` function is R code that will return the mean of a vector of numbers.

### Embedding plots

You can embed plots directly into your document using an R chunk.

```{r plotMike}
library(ggplot2)

mike <- ggplot(babynames[babynames$name == "Michael", ], aes(x = year, y = prop)) +
  geom_line(aes(color = sex)) +
  ggtitle(label = "Michael") +
  theme_bw()

mike
```