---
title: "An Example R Markdown Document"
author: "Andrew Marder"
date: "11/23/2016"
output: html_document
---

R Markdown is really cool! It allows one to combine text and code smoothly. Text is written using Markdown syntax, and code is placed inside code chunks. In RStudio, clicking the **Knit** button will generate a document that knits together the text, code, and output from executing the code chunks. Let's jump into why R Markdown is so great.

To start, let's use the `data()` function to load the `mtcars` dataset. This will set up a new data frame in memory called `mtcars`.

```{r}
data(mtcars)
```

Now that we've loaded the data into memory, we can take a look at the data. Let's use the `head()` function to view the first six rows.

```{r}
head(mtcars)
```

There are `r nrow(mtcars)` rows in the data set. Each observation corresponds to an automobile review from the 1974 Motor Trend magazine. One of R Markdown's awesome features is its ability to embed graphs. Let's create a graph examining how fuel efficiency relates to weight.

```{r}
library(ggplot2)

ggplot(mtcars, aes(y = mpg, x = 1000 * wt)) +
    geom_point() +
    ylab("Miles Per Gallon") +
    xlab("Weight (lbs)") +
    ggtitle("Fuel Efficiency as a Quadratic Function of Weight") +
    geom_smooth(method = lm, formula = y ~ poly(x, 2)) +
    theme_bw()
```

How cool is that!? Rendering this R Markdown document creates an image file of the graph above and embeds that image file in the final HTML document.

The default behavior of R Markdown is to "echo" each code chunk in the resulting document. The easiest way to remove code from the final document is the set the `echo` option to `FALSE`.

```{r, echo = FALSE}
mtcars$Transmission <- ifelse(mtcars$am, "Manual", "Automatic")

ggplot(mtcars, aes(x = qsec, fill = Transmission)) +
    geom_density(alpha = 0.5) +
    ggtitle("Where did this plot come from?") +
    xlab("1/4 Mile Time (Seconds)") +
    ylab("Density") +
    theme_bw()
```
